---
title: "Bootstrap"
author: "Meagan Lacroix"
date: "2024-11-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Compute the 95% bootstrap confidence intervals for the differences in medians of the mortality variables by armed conflict exposure
```{r}
library(here)
library(dplyr)
library(boot)
finaldata <- read.csv(here("Data", "finaldata.csv"), header = TRUE)
```

#Maternal mortality
#Interpretation: the median difference in maternal mortality between countries exposed and not exposed to armed conflict is 126.5 deaths per 100,000 (SE = 65.58). The bias corrected and accelerated boostrap 95%CI is 41.3 to 280.4 indicating the bootstrap median lies somewhere between these values. 
```{r}
data2017mat <- finaldata %>%
  filter(year == 2017 & !is.na(Mor_mat))

getmeddiffMor_mat <- function(data, indices) {
  sample_data <- data[indices, ]
  group_meds <- tapply(sample_data$Mor_mat, sample_data$armconf1, FUN = median)
  meddiff <- group_meds[2] - group_meds[1]
  return(meddiff)
}

bootoutMor_mat <- boot(data2017, statistic = getmeddiffMor_mat, strata = data2017mat$armconf1, R = 1000)
bootoutMor_mat
boot.ci(boot.out = bootoutMor_mat, conf = 0.95, type = c("basic", "perc", "bca"))
```
#Neonatal mortality
#Interpretation: the median difference in neonatal mortality between countries exposed and not exposed to armed conflict is 11.85 deaths per 100,000 (SE = 3.27). The bias corrected and accelerated boostrap 95%CI is 4.27 to 16.95 indicating the bootstrap median lies somewhere between these values.
```{r}
data2017neonat <- finaldata %>%
  filter(year == 2017 & !is.na(Mor_neonat))

getmeddiffMor_neonat <- function(data, indices) {
  sample_data <- data[indices, ]
  group_meds <- tapply(sample_data$Mor_neonat, sample_data$armconf1, FUN = median)
  meddiff <- group_meds[2] - group_meds[1]
  return(meddiff)
}

bootoutMor_neonat <- boot(data2017neonat, statistic = getmeddiffMor_neonat, strata = data2017neonat$armconf1, R = 1000)
bootoutMor_neonat
boot.ci(boot.out = bootoutMor_neonat, conf = 0.95, type = c("basic", "perc", "bca"))
```
#Under 5 mortality
#Interpretation: the median difference in under 5 mortality between countries exposed and not exposed to armed conflict is 28.9 deaths per 100,000 (SE = 9.59). The bias corrected and accelerated boostrap 95%CI is 8.93 to 45.44 indicating the bootstrap median lies somewhere between these values. 
```{r}
data2017under5 <- finaldata %>%
  filter(year == 2017 & !is.na(Mor_under5))

getmeddiffMor_under5 <- function(data, indices) {
  sample_data <- data[indices, ]
  group_meds <- tapply(sample_data$Mor_under5, sample_data$armconf1, FUN = median)
  meddiff <- group_meds[2] - group_meds[1]
  return(meddiff)
}

bootoutMor_under5 <- boot(data2017under5, statistic = getmeddiffMor_under5, strata = data2017under5$armconf1, R = 1000)
bootoutMor_under5
boot.ci(boot.out = bootoutMor_under5, conf = 0.95, type = c("basic", "perc", "bca"))
```
#Infant mortality
#Interpretation: the median difference in under infant mortality between countries exposed and not exposed to armed conflict is 19.8 deaths per 100,000 (SE = 6.06). The bias corrected and accelerated boostrap 95%CI is 6.56 to 29.18 indicating the bootstrap median lies somewhere between these values. 
```{r}
data2017inf <- finaldata %>%
  filter(year == 2017 & !is.na(Mor_inf))

getmeddiffMor_inf <- function(data, indices) {
  sample_data <- data[indices, ]
  group_meds <- tapply(sample_data$Mor_inf, sample_data$armconf1, FUN = median)
  meddiff <- group_meds[2] - group_meds[1]
  return(meddiff)
}

bootoutMor_inf <- boot(data2017inf, statistic = getmeddiffMor_inf, strata = data2017inf$armconf1, R = 1000)
bootoutMor_inf
boot.ci(boot.out = bootoutMor_inf, conf = 0.95, type = c("basic", "perc", "bca"))
```