---
title: "Exploratory Data Analysis"
author: "Meagan Lacroix"
format: pdf
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#| message = FALSE
library(here)
library(dplyr)
library(arrow)
library(janitor)
library(knitr)
library(lubridate)
library(mice)
library(modelsummary)
library(naniar)
library(tidyverse)
```

### Read in the merged data

```{r}
#| warning: false
eda_data <- read.csv(here("data", "finaldata.csv"), header = TRUE)
```

### Check names, head/tail

```{r}
names(eda_data)
head(eda_data)
tail(eda_data)
```

### Creating more consistent variable names

```{r}
eda_data <- eda_data %>%
  clean_names()
```

### Changing OECD, OECD2023, armconf1, drought, and earthquate to a factor

```{r}
eda_data <- eda_data %>%
  mutate(across(c(oecd, oecd2023, armconf1, drought, earthquake), as.factor))
```

### Summarize the variables. There is a large amount of missing data for maternal mortality. This will be investigated further

```{r}
eda_data %>% summary()
```

### Checking the names of the character variables. There are 186 countries and 17 regions

```{r}
unique(eda_data$country_name)
unique(eda_data$iso)
unique(eda_data$region)
```

### Creating histograms of the numeric variables. Total deaths is difficult to visualize because there is a large number of 0-1 deaths and a very small frequency of larger \# deaths. This will be investigated further.

```{r}
numeric_vars <- eda_data %>%
  select(where(is.numeric)) %>%  
  names() 

for (var in numeric_vars) {
  p <- ggplot(eda_data, aes(x = !!sym(var))) +  # Dynamically assign variable
    geom_histogram(binwidth = 1, fill = "blue") + 
    xlab(var) +  # Set x-axis label to the variable name
    theme_minimal()
  
  print(p)  
}
```

### 2730 (73%) rows of data had 0 battle-related deaths in the study time-frame.

```{r}

totdeath_counts <- eda_data %>%
  group_by(totdeath) %>%  
  summarize(count = n())  

print(totdeath_counts)

```

### Summary of deaths by Country.

### - Syria had the largest number of battle-related deaths

### - Sierra Leone had the most maternal, infant, and under 5 deaths

### - Pakistan had the most neonatal deaths

### - Andora, Dominica, and Marshall Islands had no maternal death data

### - Puerto Rico had no infant, neonatal, or under5 data. We may wish to exclude Puerto Rico from subsequent analyses.

```{r}
# Total deaths table with NA count by country
eda_data %>%
  group_by(country_name) %>%
  summarise(
    total_deaths = sum(totdeath, na.rm = TRUE),
    total_deaths_NA = sum(is.na(totdeath))
  ) %>%
  arrange(desc(total_deaths))

# Maternal deaths table with NA count by country
eda_data %>%
  group_by(country_name) %>%
  summarise(
    maternal_deaths = sum(mor_mat, na.rm = TRUE),
    maternal_deaths_NA = sum(is.na(mor_mat))
  ) %>%
  arrange(desc(maternal_deaths))

# Infant deaths table with NA count by country
eda_data %>%
  group_by(country_name) %>%
  summarise(
    infant_deaths = sum(mor_inf, na.rm = TRUE),
    infant_deaths_NA = sum(is.na(mor_inf))
  ) %>%
  arrange(desc(infant_deaths))

# Neonatal deaths table with NA count by country
eda_data %>%
  group_by(country_name) %>%
  summarise(
    neonatal_deaths = sum(mor_neonat, na.rm = TRUE),
    neonatal_deaths_NA = sum(is.na(mor_neonat))
  ) %>%
  arrange(desc(neonatal_deaths))

# Under-5 deaths table with NA count by country
eda_data %>%
  group_by(country_name) %>%
  summarise(
    under5_deaths = sum(mor_under5, na.rm = TRUE),
    under5_deaths_NA = sum(is.na(mor_under5))
  ) %>%
  arrange(desc(under5_deaths))
```

### Summary of deaths by year

### - The most battle-related deaths occurred in 2015

### - The most maternal, infant, neonatal, and under 5 deaths occurred in 2000

### - There is no maternal death data in 2018/2019

```{r}
# Total deaths by year with NA count
eda_data %>%
  group_by(year) %>%
  summarise(
    total_deaths = sum(totdeath, na.rm = TRUE),
    total_deaths_NA = sum(is.na(totdeath))
  ) %>%
  arrange(desc(total_deaths))

# Maternal deaths by year with NA count
eda_data %>%
  group_by(year) %>%
  summarise(
    maternal_deaths = sum(mor_mat, na.rm = TRUE),
    maternal_deaths_NA = sum(is.na(mor_mat))
  ) %>%
  arrange(desc(maternal_deaths))

# Infant deaths by year with NA count
eda_data %>%
  group_by(year) %>%
  summarise(
    infant_deaths = sum(mor_inf, na.rm = TRUE),
    infant_deaths_NA = sum(is.na(mor_inf))
  ) %>%
  arrange(desc(infant_deaths))

# Neonatal deaths by year with NA count
eda_data %>%
  group_by(year) %>%
  summarise(
    neonatal_deaths = sum(mor_neonat, na.rm = TRUE),
    neonatal_deaths_NA = sum(is.na(mor_neonat))
  ) %>%
  arrange(desc(neonatal_deaths))

# Under-5 deaths by year with NA count
eda_data %>%
  group_by(year) %>%
  summarise(
    under5_deaths = sum(mor_under5, na.rm = TRUE),
    under5_deaths_NA = sum(is.na(mor_under5))
  ) %>%
  arrange(desc(under5_deaths))

```

### Visualizing the amount of missing data across all variables. There is a high level of missingness for maternal mortality (11% missing). All other variables have less than 5% missing data.

```{r}
vis_miss(eda_data)
```

### Further visualizing the missingness by year. 2018/2019 has the most missing data for maternal mortality.

```{r}
gg_miss_var(eda_data, facet = year)

```

### Checking the correlations of the numeric variables

```{r}
numeric_data <- eda_data[, sapply(eda_data, is.numeric)]

numeric_data <- numeric_data %>%
  select(-year)

cor(numeric_data, use = "complete.obs")
```

### Generating boxplots to show the relationship between the mortality variables and the binary variables.

-   OECD countries have far fewer battle-related deaths as well as maternal, neonatal, infant, and under 5 deaths

-   Median death count for all mortality variables is higher when there is an armed conflict

-   Median death count for all mortality variables is only slightly higher when there is an earthquake

-   Median death count for all mortality variables is higher when there is a drought

```{r}

generate_boxplots <- function(data, binary_vars) {
  # Reshape the data to long format for the death variables
  data_long <- data %>%
    pivot_longer(cols = c(mor_inf, mor_mat, mor_under5, mor_neonat, totdeath),
                 names_to = "death_type", values_to = "death_value")

  # Loop through each binary variable
  for (binary_var in binary_vars) {
    # Create boxplot for the current binary variable
    p <- ggplot(data_long, aes_string(x = binary_var, y = "death_value"))+
      geom_boxplot() +
      facet_wrap(~ death_type, scales = "free") +
      labs(x = paste(binary_var, "(0 = No, 1 = Yes)"), y = "Death Count",
           title = paste("Boxplots of Death Variables by", binary_var)) +
      theme_minimal()
    
    # Print the plot
    print(p)
  }
}


binary_variables <- c("oecd", "oecd2023", "armconf1", "earthquake", "drought")   

generate_boxplots(eda_data, binary_variables)
```

### Creating scatterplots to visualize the relationship between the mortality variables and the other continuous variables.

-   There is a negative relationship between all mortality variables and population density, urban residence, and male education

-   There is a positive relationship between the mortality variables and age dependency and temperature

-   The relationship between the mortality variables and rainfall is less clear and warrants further exploration

```{r}
generate_scatterplots <- function(data, continuous_vars) {
  # Reshape the data to long format for the death variables
  data_long <- data %>%
    pivot_longer(cols = c(mor_inf, mor_mat, mor_under5, mor_neonat, totdeath),
                 names_to = "death_type", values_to = "death_value")

  # Loop through each binary variable
  for (continuous_vars in continuous_vars) {
    # Create boxplot for the current binary variable
    p <- ggplot(data_long, aes_string(x = continuous_vars, y = "death_value"))+
      geom_point(alpha=.5) +
      geom_smooth(method = "lm", se = FALSE, color = "blue") +
      facet_wrap(~ death_type, scales = "free") +
      labs(x = paste(continuous_vars), y = "Death Count",
           title = paste("Scatterplot of Mortality Variables by", continuous_vars)) +
      theme_minimal()
    
    # Print the plot
    print(p)
  }
}


continuous_vars <- c("popdens", "urban", "agedep", "male_edu", "temp", "rainfall1000")   

generate_scatterplots(eda_data, continuous_vars)
```
